Clustering time series based on probability distributions across temporal granularities

### 2021-10-26 main.Rmd
introduction changed with new structure
intro updated

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<<n$) groups such that homogeneous time-series are grouped together based on a certain similarity measure. The appropriate similarity measure or the distance approach is determined by the time-series characteristics, duration of time-series, representation method, and, of course, the goal of clustering time-series to a significant level. The three main approaches to time series clustering (@liao2005clustering) are algorithms that work directly with distances or raw data points, either in the time or frequency domain (distance-based), with features extracted from raw data (feature-based), or indirectly with models built from the raw data (model-based). The performance of distance-based approaches depends greatly on the particular distance metric used. Defining an adequate distance measure for the raw time series can be a challenging task since it has to consider noise, different length of time series, asynchronous time series, different scales, missing values. Distance-based similarity measures described in the literature include
Euclidean,  Pearson’s correlation coefficient and related distances [124], Dynamic Time Warping [118,119], Autocorrelation, Short timeseries distance [44], Piecewise normalization [125], cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances.
Euclidean distance and DTW are the most common methods for similarity measure in time-series clustering. A research has shown that, in terms of time-series classification accuracy, the Euclidean distance is surprisingly competitive [145], however, DTW also has its strength in similarity measurements which cannot be declined. While these methods have been quite effective in clustering moderate-length time series, there are some well-recognized drawbacks of the existing approaches. Euclidean distance is the most common method for discerning similarity in time series clustering, and it requires that the time series being compared are of exactly the same dimensionality (length). DTW can assist with clustering of different length time series if there is no missing data (Ratanamahatana and Keogh, 2005) and in any case it requires quadratic computation.

We are motivated by the residential smart meter data. These long time series are asynchronous, with varying time lengths for different houses and missing observations and characterised by noisy and patchy behavior that can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Choosing probability distributions instead of raw data seems to be a natural way to analyze these types of data sets. Hence this paper proposes a distance metric based on Jensen-Shannon distances between probability distributions across significant cyclic granularities. Cyclic temporal granularities, which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. Since cyclic granularities are considered instead of linear granularities, the resulting clusters are expected to group customers that have similar repetitive behaviors. Below we list some of the advantages of our approach, which will be explained in more detail along with empirical results in later sections.

- When the time series is long (high dimensionality), some clustering algorithms become intractable. By applying dimensionality reduction via characterising probability distributions, we are able to cluster long length time series efficiently;

- When the clustering algorithm is based on a distance metric (i.e. Euclidean distance), it cannot handle time series with missing data or of different lengths if actual points are used as inputs. However, by
using probability distribution, we can trim the tails and bypass that problem.

- We recognize that most clustering algorithms provide profiles across only one temporal granularity, but this approach provides a wider approach to the issue, seeking to group observations with similar shapes over all important cyclic granularities. Since cyclic granularities are considered instead of linear granularities, clustering would group customers that have similar repetitive behavior across more than one cyclic granularities across which patterns are expected to be significant.

- It is reasonable to define a time series according to its degree of trend and seasonality and consider these features while clustering them. These characteristics of the time series are lost or handled independently by considering probability distributions (trend is lost) across categories of cyclic granularities (by independently modeling all seasonal fluctuations), and so there is no need to de-trend or de-seasonalize the data before conducting the clustering method. For example, there is no need to omit holiday or weekend patterns for similar reasons.

@@

<!-- significance of load clustering -->
Large spatio-temporal data sets, both from open and administrative sources, offer up a world of possibilities for research. <!--Beyea (2010) has pointed out, there has been little exploration of the full potential of these data bases and their benefits can reach beyond the original intentions for collecting these data.--> One such data sets for Australia is the Smart Grid, Smart City (SGSC) project (2010–2014) available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee)<!--and Data61 CSIRO-->. The project provides half-hourly data of over 13,000 household electricity smart meters distributed unevenly from October 2011 to March 2014. <!---The massive amount of data generated in such projects could be overwhelming for analysis.--> Raw data of these asynchronous time series can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Electricity utilities can utilize the smart meter usage patterns to develop targeted tariffs for individual groups and alleviate the problem of volatility in production by capitalizing on the flexibility of consumers.

<!-- Sources of variation in large data sets -->
<!-- The enormous quantity of data provides for greater individual level clarity and analysis. However, -->
<!-- due to the growing variety of consumers,  -->

\noindent Larger data sets include greater uncertainty about customer behavior due to growing variety of customers. Households are distributed geographically, and have different demographic properties such as the existence of solar panels, central heating or air conditioning. Multiple temporal dependencies define the behavioral patterns, which vary from customer to customer. Some families, for example, use a dryer to dry their clothing, while others hang them to dry on a line. This may be reflected in their weekly profile. They may have monthly variations where some customers are more prone to use air conditioners or heaters than others despite the existence of comparable electrical equipment and being subjected to similar weather conditions. The variations in behavior may occur on a regular basis, with some consumers being night owls and others being morning larks. Day-off energy usage may vary depending on whether consumers choose to remain at home or engage in outdoor activities. These various causes of variations in energy behavior, ranging from age, lifestyle, and family composition to building characteristics, weather, presence of various electrical equipments and others, make the problem of effectively segmenting consumers into similar energy behavior a particularly intriguing one.

<!-- unsupervised clustering -->
\noindent This problem becomes more difficult when there is no other information on the customers besides their time series of energy usage. It is likely that such additional information is not available in order to protect the customers' confidentiality. Furthermore, it is not guaranteed that energy providers would always update customer profiles, such as property characteristics, in a timely manner whenever they change.
So this study does not attempt to explain why consumption differs. Instead, the work looks at how much energy-use heterogeneity can be uncovered in smart meter data and what are some of the most typical electricity use patterns by simply using the time series.

@@

A typical clustering technique includes the following steps: (a) establishing distance (dissimilarity) and similarity through feature or model extraction and selection; and (b) selecting the clustering algorithm design. Distance measures could be time-domain based or frequency-domain (fast Fourier transform). A model-based clustering works by transforming the series into other other objects such as structure or set of parameters which can be more easily characterised and clustered (@Motlagh2019-yj). [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. The essential temporal characteristics of the curves are defined or extracted using feature-based clustering. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. None of the $34$ articles chosen for their emphasis use Australian smart meter data. The most often used algorithm is K-Means. Using K-Means without considering time series structure or correlation results in inefficient clusters. Principal Component Analysis (PCA) or Self-Organizing Maps (SOM) eliminate correlation patterns and decrease feature space, but lose interpretability. To reduce dimensionality, several studies use principal component analysis or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). Other algorithms utilised in the literature include k-means variants, hierarchical approaches, and greedy k-medoids. Time series data, such as smart metre data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics using Fourier transformation, which converts data from time to frequency and then uses K-Means to cluster by greatest frequency .

@@

The remainder of the paper is organized as follows: Section&nbsp;\ref{sec:methodology} provides the clustering methodology introducing the features and distance metrics.
Section&nbsp;\ref{sec:validation} shows data designs to validate our methods and draw comparisons against several methods. Section&nbsp;\ref{sec:application} discusses the application of the method to a subset of the real data. Finally, we summarize our results and discuss possible future directions in Section&nbsp;\ref{sec:discussion}.

>>>>

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance strategy to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based) (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilised. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used Distance-based similarity measures as suggested by a review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping, Autocorrelation, Short time series distance, Piecewise regularisation, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering). Euclidean distance and DTW are often used in time series clustering. When it comes to time-series clustering accuracy, the Euclidean distance beats DTW, but DTW has its own advantages (@corradini2001dynamic).Euclidean distance requires time series of equal length. while DTW can assist cluster time series of varying lengths (@ratanamahatana2005multimedia), only if there are no missing observations.

We are motivated by the residential smart meter data. These long time series are asynchronous, with varying time lengths for different houses and missing observations and characterised by noisy and patchy behavior that can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Choosing probability distributions instead of raw data seems to be a natural way to analyze these types of data sets. Hence this paper proposes a distance metric based on Jensen-Shannon distances between probability distributions across significant cyclic granularities. Cyclic temporal granularities, which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. Since cyclic granularities are considered instead of linear granularities, the resulting clusters are expected to group customers that have similar repetitive behaviors.  Below are some of the benefits of our method, which will be detailed in further depth in subsequent sections.

- Some clustering algorithms become problematic with the very high dimensionality of the time series resulting from the frequency at which they are recorded and the length of time for which they are observed. We can efficiently cluster long length time series by reducing dimensionality by characterising through probability distributions;

- By utilising Jensen-Shannon distances, we are evaluating the distance between two distributions rather than raw data, which is less susceptible to missing observations and outliers compared to other traditional distance measures;

- While most clustering algorithms produce clusters similar across just one temporal granularity, this technique takes a broader approach to the problem, attempting to group observations with similar forms across all key cyclic granularities. Because cyclic granularities are used rather than linear granularities, clustering would group consumers who exhibit similar repeating behaviour over many cyclic granularities where patterns are predicted to be important.

- It is reasonable to define a time series based on its degree of trend and seasonality and to take these characteristics into account while clustering it. The change in data structure by considering probability distributions across cyclic granularities ensures there is no trend and seasonal fluctuations are handled separately. Thus there is no need to de-trend or de-seasonalize the data prior to performing the clustering method. For similar reasons, there is no need to exclude holiday or weekend routines.

@@

<!-- significance of load clustering -->
Large spatio-temporal data sets, both from open and administrative sources, offer up a world of possibilities for research. <!--Beyea (2010) has pointed out, there has been little exploration of the full potential of these data bases and their benefits can reach beyond the original intentions for collecting these data.--> One such data sets for Australia is the Smart Grid, Smart City (SGSC) project (2010–2014) available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee)<!--and Data61 CSIRO-->. The project provides half-hourly data of over 13,000 household electricity smart meters distributed unevenly from October 2011 to March 2014. <!---The massive amount of data generated in such projects could be overwhelming for analysis.Raw data of these asynchronous time series can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Electricity utilities can utilize the smart meter usage patterns to develop targeted tariffs for individual groups and alleviate the problem of volatility in production by capitalizing on the flexibility of consumers. --><!-- Sources of variation in large data sets --><!-- The enormous quantity of data provides for greater individual level clarity and analysis. However, -->
<!-- due to the growing variety of consumers,  -->. Larger data sets include greater uncertainty about customer behavior due to growing variety of customers. Households vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioural patterns differ amongst customers due to many temporal dependencies. Some households use a dryer, while others dry their clothes on a line. Their weekly profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Day-off energy use varies depending on whether customers stay home or go outside. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behaviour a fascinating one. This challenge is worsened when all we know about our consumers is their energy use history (@Ushakova2020-rl). To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner. Thus, there is a growing need to have research that examines how much energy usage heterogeneity can be found in smart meter data and what are some of the most common power consumption patterns, rather than explaining why consumption differs.

@@

A multitude of papers have emerged around smart meter time series clustering for deepening our knowledge of consumption patterns. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. None of the $34$ articles chosen for their emphasis use Australian smart meter data. The most often used algorithm is K-Means. Using K-Means without considering time series structure or correlation results in inefficient clusters. Principal Component Analysis (PCA) or Self-Organizing Maps (SOM) eliminate correlation patterns and decrease feature space, but lose interpretability. To reduce dimensionality, several studies use principal component analysis or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). Other algorithms utilised in the literature include k-means variants, hierarchical approaches, and greedy k-medoids. Time series data, such as smart metre data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics using Fourier transformation, which converts data from time to frequency and then uses K-Means to cluster by greatest frequency . @Motlagh2019-yj suggests that the time feature extraction islimited by the type of noisy, patchy, and unequal time-series common in residential datasets and addresses model-based clustering  by transforming the series into other other objects such as structure or set of parameters which can be more easily characterised and clustered. [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp disucusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions.

\noindent Given the limitations of the similarity measures in dealing with large volumes of this complicated time series data, we present a similarity measure based on probability distributions that seems to be a more organic option for coping with time series data with aforementioned characteristics. The remainder of the paper is organized as follows: Section&nbsp;\ref{sec:methodology} provides the clustering methodology introducing the features and distance metrics. Section&nbsp;\ref{sec:validation} shows data designs to validate our methods and draw comparisons against several methods.Section&nbsp;\ref{sec:application} discusses the application of the method to a subset of the real data. Finally, we summarize our results and discuss possible future directions in Section&nbsp;\ref{sec:discussion}.

<!-- A typical clustering technique includes the following steps: (a) establishing distance (dissimilarity) and similarity through feature or model extraction and selection; and (b) selecting the clustering algorithm design. Distance measures could be time-domain based or frequency-domain (fast Fourier transform). -->

<!-- A model-based clustering works by transforming the series into other other objects such as structure or set of parameters which can be more easily characterised and clustered (@Motlagh2019-yj). [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. The essential temporal characteristics of the curves are defined or extracted using feature-based clustering. -->

@@
